Optimality and Heuristics for Finite Element Integration

### 2015-10-28 TOMS_Paper/assembly.tex
Trying to fix sharing elimination, still...UFF

In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (potentially) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned. In the following, we assume outer product loop nests of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$.

@@

\begin{enumerate}
\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed. This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$). Hoisting and capturing these sub-expressions may make the operation count of outer loops lower than in point 1), but the gain (if any) will be negligible (note: the operation count in the innermost loop does not vary).
\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in a scenario in which loops degenerate, i.e. with iteration spaces of size 1).
\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have multiple implications. We distinguish two cases:
\begin{enumerate}
\item The factorization occurs before points 1) or 2). This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) simply reduces to cases 1) or 2) (for example, in the case of forms with scalar coefficients).
\item The factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. If the loop trip count is extremely low, however, this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$. It is true that factorizing $v_i$ and $w_i$ would expose $L_i$-independent sub-expressions, but this would actually result in increased operation count if, for instance, the size of $L_i$ were 1.
\end{enumerate}
\end{enumerate}

>>>>

In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned. In the following, we assume outer product loop nests of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$.

@@

\begin{enumerate}
\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed. This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$). Hoisting and capturing these sub-expressions may make the operation count of outer loops lower than in point 1), but the gain (if any) will be negligible (note: the operation count in the innermost loop does not vary).
\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in a scenario in which loops degenerate, i.e. with iteration spaces of size 1).
\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have multiple implications. We distinguish two cases:
\begin{enumerate}
\item Factorization occurs in place of points 1) or 2). This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) simply reduces to cases 1) or 2) (for example, in the case of forms with scalar coefficients).
\item Factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. If the loop trip count is extremely low, however, this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$. It is true that factorizing $v_i$ and $w_i$ would expose $L_i$-independent sub-expressions, but this would actually result in increased operation count if, for instance, the size of $L_i$ were 1.
\end{enumerate}
\end{enumerate}
